The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-08 dictionary.tex
Added r : V -> V \otimes V to finite QCS
Working on automorphism groups

In this section we define \emph{finite quasicharacter sheaves}, a further refinement
of bounded quasicharacter sheaves; finite quasicharacter sheaves will play a role in Section \ref{sec:restriction}.

The notion of a finite quasicharacter sheaf is informed
by a certain perspective on local systems on connected groups.
Recall that every geometric point $\bg$ on a connected scheme $G$
determines an equivalence between the categories of $\ell$-adic local systems on $G$
and $\ell$-adic representations of the \'etale fundamental group $\pi_1(G, \bg)$
\citelist{\cite{SGA5}*{VI,Prop. 1.2.5} \cite{deligne:80a}*{1.1.1}}.
 Since $\pi_1(G, \bg)$ is a projective limit of automorphism groups of finite \'etale covers,
 a general local system corresponds to a projective limit of characters of such groups.
 But some local systems can be described in terms of a single cover $f: H \to G$
 and a representation $\psi: \ker f \to \Aut(V)$;
 two such descriptions yield isomorphic local systems if and only if they are related by a weak
 isomorphism in $C(G)$.  While the description of local systems in terms of $\pi_1(G, \bg)$ does
 not adapt well to non-connected $G$, a perspective using finite \'etale covers does.

The definition of finite quasicharacter sheaves will require several steps,
beginning with the localization of a category introduced here.
Let us say that a morphism $f : H\to G$ of smooth commutative group schemes over $\Fq$ is a \emph{discrete isogeny}
if it is finite, surjective, \'etale and the action of $\Gal(\bFq/\Fq)$ on the \'etale group
scheme $\ker f = f^{-1}({\bar e)} = \Aut(f)$ is trivial.
Consider the category $C(G)$ whose objects consist of triples $(f,\psi,V)$, where
\begin{itemize}
\item $f : H\to G$ is a discrete isogeny of smooth commutative group schemes over $\Fq$,
\item $\psi : \ker f\to \Aut(V)$ is a representation on a finite-dimensional $\EE$-vector space $V$.
\end{itemize}
A morphism $(f,\psi,V) \to (f',\psi',V')$ in $C(G)$ is a pair $(\gamma,T)$, where
\begin{itemize}
\item $\gamma : H' \to H$ is a morphism of group schemes such that $f' = f\circ \gamma$,
\item $T : V\to V'$ is a linear transformation, equivariant for the action of
$\ker f'$ on $V'$ by $\psi'$ and the action of $\ker f'$ on $V$ by $\psi \circ \alpha$.
\[
\begin{tikzcd}
V \arrow{d}{T} & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} & \arrow[equal]{d} G\\
V' & \Aut(V') & \arrow[swap]{l}{\psi'} \arrow[swap]{u}{\gamma\vert_{\ker f'}} \ker f' \arrow[hook]{r} & \arrow[swap]{u}{\gamma} H' \arrow{r}{f'} & G
\end{tikzcd}
\]
\end{itemize}

We put a tensor category structure on $C(G)$ as follows.
Let $(f,\psi,V)$ and $(f',\psi',V')$ be objects in $C(G)$.
Write $f\times_G f' : H\times_G H' \to G$ for the product of $f$ and $f'$
in the category of isogenies to $G$. We define an external tensor product
$\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes V')$ and a external sum
$\psi \oplus\psi' :  \ker(f\times_G f') \to \Aut(V\oplus V')$
using the canonical isomorphism $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$.
We may thus define $(f,\psi,V)\otimes(f',\psi',V') \ceq (f\times_G f' ,\psi\otimes\psi', V\otimes V')$ and
$(f,\psi,V)\oplus(f',\psi',V')\ceq (f\times_G f' ,\psi\oplus\psi', V\oplus V')$. To define duals in $C(G)$,
let $V^\vee \ceq \Hom_\text{vec}(V,\EE)$ and $\psi^\vee : \ker f \to \Aut(V^\vee)$ be
the dual representation.  We may then set $(f,\psi,V)^\vee \ceq (f,\psi^\vee, V^\vee)$.
These operations give $C(G)$ the structure of a tensor category.

Now we localize $C(G)$.
Let us say that $(\gamma,T)$ in $C(G)$ is a \emph{weak isomorphism} if $T$ is an isomorphism of vector spaces.  
Let $W$ be the family of weak isomorphism and
et $C(G)[W^{-1}]$ be the localization of $C(G)$ with respect to $W$, 
in the sense of \cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.1}.
Since weak isomorphisms in $C(G)$ form a \emph{right multiplicative system}, in the sense of \cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.5}, it follows from \cite{kashiwara-schapira:CatgoriesSheaves}*{Thm 7.1.16} that morphisms in $C(G)[W^{-1}]$ are simple to describe: 
a morphism in $C(G)[W^{-1}]$ is an equivalence class of diagrams
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,V) \arrow{r}{(\gamma,T)} & (f',\psi',V') & \arrow[swap]{l}{(\gamma_1,T_1)}[swap]{\text{w-iso}} (f_1,\psi_1,V_1)
   \end{tikzcd}
\]
 where $(\gamma,T) : (f,\psi,V) \to (f',\psi',V')$ is any map in $C(G)$
 and $(\gamma_1, T_1) :  (f',\psi',V') \to (f_1,\psi_1,V_1)$ is a weak isomorphism in $C(G)$;
for the equivalence relation on these diagrams, see \cite{kashiwara-schapira:CatgoriesSheaves}*{(7.1.3)}.

It is straightforward to verify that the subcategory $C_1(G)$ of $C(G)$ consisting of $(f,\psi,V)$ for which $\psi$ is one-dimensional, is a full, rigid monoidal subcategory of $C(G)$ and further, that the family $W_1$ of weak isomorphisms in $C_1(G)$ is a right multiplicative system.

@@

Trivial morphisms in $\rfQC(G)$ are defined as follows. Let $(f,\psi,V)$ and $(f_1,\psi_1,V_1)$ be finite quasicharacter sheaves on $G$.  
Form $(f\times_G f_1,1\otimes \psi_1, V\otimes V_1)$ in $C_1(G)$. 
Let $0 : V \to V\otimes V_1$ be the trivial linear transformation. 
Choose $0\ne v \in V$ and let $T_v: V_1 \to V\otimes V_1$ be the isomorphism $T_v(v_1) = v\otimes v_1$.  
Let $\pi : H\times_G H_1 \to H$ and $\pi_1 : H\times_G H_1 \to H_1$ be the natural projections.
Then $(\pi,0) : (f,\psi, V) \to  (f\times_G f_1,1\otimes \psi_1, V\otimes V_1)$ is a morphism in $C_1(G)$ and $(\pi_1,T_v) : (f_1,\psi_1,V_1) \to (f\times_G f_1,1\otimes \psi_1, V\otimes V_1)$ is a weak isomorphism in $C_1(G)$.
%
The morphism in $\rfQC(G)$ represented by
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,V) \arrow{r}{(\pi,0)} & (f\times_G f_1,1\otimes \psi_1, V\otimes V_1) & \arrow[swap]{l}{(\pi_1,T_v)}[swap]{\text{w-iso}}  (f_1,\psi_1,V_1),
   \end{tikzcd}
\]
does not depend on the choice of $0\ne v\in V$; this defines the {\em trivial morphism} $0: (f,\psi,V) \to (f_1,\psi_1,V_1)$ in $\rfQC(G)$.

\begin{proposition}\label{prop:autornaught}
Every morphism in $\rfQC(G)$ is either an isomorphism or trivial.
\todo{I may add something here about $\Aut_{\rfQC(G)}(f,\psi,V)$.}
\end{proposition}

\begin{proof}
Let $(f,\psi,V)$ and $(f_1,\psi_1,V_1)$ be finite quasicharacter sheaves on $G$. 
Let
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,V) \arrow{r}{(\gamma,T)} & (f',\psi',V') & \arrow[swap]{l}{(\gamma_1,T_1)}[swap]{\text{w-iso}}  (f_1,\psi_1,V_1)
   \end{tikzcd}
\]
represent a morphism $(f,\psi,V)  \to (f_1,\psi_1,V_1)$ in $\rfQC(G)$;
note that, by definition, $(\gamma_1 , T_1) : (f_1,\psi_1,V_1) \to (f',\psi',V')$ is a weak isomorphism.
Since $V$, $V'$ and $V_1$ are one-dimensional $\EE$-vector spaces, either $T : V \to V'$ is an isomorphism or trivial.

>>>>

In this section we define \emph{finite quasicharacter sheaves}, 
a notion informed by a certain perspective on local systems on connected groups.
Finite quasicharacter sheaves will play a role in Section \ref{sec:restriction}.

%The definition of finite quasicharacter sheaves will require several steps, beginning with the localization of a category introduced here.
Consider the category $C(G)$ whose objects consist of triples $(f,\psi,r)$, where:
\begin{itemize}
\item $f : H\to G$ is a \emph{discrete isogeny} (a morphism of smooth commutative group schemes over $\Fq$ which is finite, surjective and \'etale and for which the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $\ker f = f^{-1}({\bar e)} = \Aut(f)$ is trivial);
\item $\psi : \Aut(f)\to \Aut(V)$ is a finite-dimensional $\EE$-representation;
\item $r : V \to V\otimes V$ is an injective linear transformation.
\end{itemize}
A morphism $(f,\psi,r) \to (f',\psi',r)$ in $C(G)$ is a pair $(\gamma,T)$, where:
\begin{itemize}
\item $\gamma : H' \to H$ is a morphism of group schemes such that $f\circ \gamma = f'$;
\item $T: V\to V'$ is a linear transformation, equivariant for the action of
$\Aut(f')$ on $V'$ by $\psi'$ and the action of $\Aut(f')$ on $V$ by $\psi \circ \gamma$,
\end{itemize}
such that
\[
\begin{tikzcd}
\arrow{d}[swap]{T\otimes T} V\otimes V & \arrow{l}{r} V \arrow{d}{T} 
%& \Aut(V) & \arrow[swap]{l}{\psi} \Aut(f) \arrow{r} & H \arrow{r}{f} & \arrow[equal]{d} G
\\
V'\otimes V' & \arrow{l}{r'} V' 
%& \Aut(V') & \arrow[swap]{l}{\psi'} \arrow[swap]{u}{} \Aut(f') \arrow{r} & \arrow[swap]{u}{\gamma} H' \arrow{r}{f'} & G
\end{tikzcd}
\]
commutes.

We define a tensor product in $C(G)$ as follows.
Let $(f,\psi,r)$ and $(f',\psi',r')$ be objects in $C(G)$.
Write $f\times_G f' : H\times_G H' \to G$ for the product of $f$ and $f'$
in the category of isogenies to $G$. We define an external tensor product
$\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes V')$ and a external sum
$\psi \oplus\psi' :  \ker(f\times_G f') \to \Aut(V\oplus V')$
using the canonical isomorphism $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$.
We may thus define $(f,\psi,r)\otimes(f',\psi',r') \ceq (f\times_G f' ,\psi\otimes\psi', r\otimes r')$ and
$(f,\psi,r)\oplus(f',\psi',r')\ceq (f\times_G f' ,\psi\oplus\psi', r\oplus r')$. 
To define duals in $C(G)$,
let $V^\vee \ceq \Hom_\text{vec}(V,\EE)$ and $\psi^\vee : \ker f \to \Aut(V^\vee)$ be
the dual representation.  We may then set $(f,\psi,r)^\vee \ceq (f,\psi^\vee, r^\vee)$ 
where $r^\vee : V^\vee \to V^\vee \otimes V^\vee$ is the adjoint of $r$.
These operations give $C(G)$ the structure of a tensor category.

Now we localize $C(G)$.
Let us say that $(\gamma,T)$ in $C(G)$ is a \emph{weak isomorphism} if $T$ is an isomorphism of vector spaces.  
Let $W$ be the family of weak isomorphism and
et $C(G)[W^{-1}]$ be the localization of $C(G)$ with respect to $W$, 
in the sense of \cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.1}.
Since weak isomorphisms in $C(G)$ form a \emph{right multiplicative system}, in the sense of \cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.5}, it follows from \cite{kashiwara-schapira:CatgoriesSheaves}*{Thm 7.1.16} that morphisms in $C(G)[W^{-1}]$ are simple to describe: 
a morphism in $C(G)[W^{-1}]$ is an equivalence class of diagrams
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,r) \arrow{r}{(\gamma,T)} & (f',\psi',r') & \arrow[swap]{l}{(\gamma_1,T_1)}[swap]{\text{w-iso}} (f_1,\psi_1,V_1)
   \end{tikzcd}
\]
 where $(\gamma,T) : (f,\psi,r) \to (f',\psi',r')$ is any map in $C(G)$
 and $(\gamma_1, T_1) :  (f',\psi',r') \to (f_1,\psi_1,r_1)$ is a weak isomorphism in $C(G)$;
for the equivalence relation on these diagrams, see \cite{kashiwara-schapira:CatgoriesSheaves}*{(7.1.3)}.

It is straightforward to verify that the subcategory $C_1(G)$ of $C(G)$ consisting of $(f,\psi,r)$ for which $\psi$ is one-dimensional, is a full, rigid monoidal subcategory of $C(G)$ and further, that the family $W_1$ of weak isomorphisms in $C_1(G)$ is a right multiplicative system.

@@

\begin{remark}
Recall that every geometric point $\bg$ on a connected scheme $G$
determines an equivalence between the categories of $\ell$-adic local systems on $G$
and $\ell$-adic representations of the \'etale fundamental group $\pi_1(G, \bg)$
\citelist{\cite{SGA5}*{VI,Prop. 1.2.5} \cite{deligne:80a}*{1.1.1}}.
 Since $\pi_1(G, \bg)$ is a projective limit of automorphism groups of finite \'etale covers,
 a general local system corresponds to a projective limit of characters of such groups.
 But some local systems can be described in terms of a single cover $f: H \to G$
 and a representation $\psi: \ker f \to \Aut(V)$;
 two such descriptions yield isomorphic local systems if and only if they are related by a weak
 isomorphism in $C(G)$.  While the description of local systems in terms of $\pi_1(G, \bg)$ does
 not adapt well to non-connected $G$, a perspective using finite \'etale covers does.
\end{remark}

Later in the paper we will make a study of isomorphism classes of objects in $\rfQC{G}$. Here we focus on morphisms in this category.

\begin{proposition}\label{prop:autornaught}
Every morphism in $\rfQC(G)$ is either an isomorphism or trivial.
Moreover, for every finite quasicharacter sheaf $(f,\psi,r)$, the group $\Aut_{\rfQC(G)}(f,\psi,r)$ is trivial.
\end{proposition}

\begin{proof}
Trivial morphisms in $\rfQC(G)$ are defined as follows. Let $(f,\psi,r)$ and $(f_1,\psi_1,r_1)$ be finite quasicharacter sheaves on $G$.  
Form $(f\times_G f_1,1\otimes \psi_1, r\otimes r_1)$ in $C_1(G)$. 
Let $0 : V \to V\otimes V_1$ be the trivial linear transformation. 
Choose $0\ne v \in V$ and let $T_v: V_1 \to V\otimes V_1$ be the isomorphism $T_v(v_1) = v\otimes v_1$.  
Let $\pi : H\times_G H_1 \to H$ and $\pi_1 : H\times_G H_1 \to H_1$ be the natural projections.
Then $(\pi,0) : (f,\psi, V) \to  (f\times_G f_1,1\otimes \psi_1, V\otimes V_1)$ is a morphism in $C_1(G)$ and $(\pi_1,T_v) : (f_1,\psi_1,V_1) \to (f\times_G f_1,1\otimes \psi_1, V\otimes V_1)$ is a weak isomorphism in $C_1(G)$.
%
The morphism in $\rfQC(G)$ represented by
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,V) \arrow{r}{(\pi,0)} & (f\times_G f_1,1\otimes \psi_1, V\otimes V_1) & \arrow[swap]{l}{(\pi_1,T_v)}[swap]{\text{w-iso}}  (f_1,\psi_1,V_1),
   \end{tikzcd}
\]
does not depend on the choice of $0\ne v\in V$; this defines the {\em trivial morphism} $0: (f,\psi,V) \to (f_1,\psi_1,V_1)$ in $\rfQC(G)$.

Let $(f,\psi,V)$ and $(f_1,\psi_1,V_1)$ be finite quasicharacter sheaves on $G$. 
Let
\[
\begin{tikzcd}[column sep=30]
   (f,\psi,V) \arrow{r}{(\gamma,T)} & (f',\psi',V') & \arrow[swap]{l}{(\gamma_1,T_1)}[swap]{\text{w-iso}}  (f_1,\psi_1,V_1)
   \end{tikzcd}
\]
represent a morphism $(f,\psi,V)  \to (f_1,\psi_1,V_1)$ in $\rfQC(G)$;
note that, by definition, $(\gamma_1 , T_1) : (f_1,\psi_1,V_1) \to (f',\psi',V')$ is a weak isomorphism.
Since $V$, $V'$ and $V_1$ are one-dimensional $\EE$-vector spaces, either $T : V \to V'$ is an isomorphism or trivial.
